---
title: "Lab3"
output: html_document
date: "2022-11-27"
---


```{r}
options(warn=-1)
library(markovchain)
```

# Problem 1

In the first part, we will estimate the probability that a random digit sequence of length n contains the TLN (consider the cases n = 100, n = 200, n = 1000).

## 1

1.  Estimate numerically the probability ˆpn of the event that your TLN occurs in a random digit sequence d1d2d3 ...dn. Hint: Such a sequence can be generated with R command sample(0:9, n, replace=T); you will need to generate a sample of such sequences of sufficiently large size N

```{r}
count_ones <- function(test, desired="003"){
    test = paste(test, collapse = "")
    result =lengths(regmatches(test, gregexpr(desired, test)))
    if (result > 1){
      result = 1
    }
    return(result)
}
```

```{r}
 get_N_results <- function(N, n, desired="003"){
    result = sample(0, size=N, r=T)
    for (i in 1:N){
      result[i] =  count_ones(sample(0:9, n, replace=T), desired)
    }
    return  (result)
 }
 
```

```{r}
desired = "003"

res_100 = get_N_results(1e5, 1e2, desired)
res_200 = get_N_results(1e5, 2e2, desired)
res_1000 = get_N_results(1e5, 1e3, desired)

```

```{r}
cat("Probability for n = 100 after N = 1e5 iterations: ", mean(res_100), "\n")
cat("Probability for n = 200 after N = 1e5 iterations: ", mean(res_200), "\n")
cat("Probability for n = 1000 after N = 1e5 iterations: ", mean(res_1000), "\n")

```

## 2

Identify the the Markov chain structure with four states S0,S1,S2,S3 in this sequence with Sk denoting the number of correct last digits (eg., for the team id number 028 these states will be S0 ="\*", S1 ="0", S2 ="02", S3 ="028"). Determine the transition probabilities matrix P and find the limiting probability pn for the state "028". Compare with the result obtained in part 1.

Markov chain for 003

S1 = "\*", S2 = 0, S3 = 00, S4 = 003

$$P=\begin{pmatrix}
0.9 & 0.1 & 0 & 0
\\0.9 & 0 & 0.1 & 0
\\0.8 & 0 & 0.1 & 0.1
\\0.9 & 0.1 & 0 & 0
\end{pmatrix}\\ $$

System of equations: $$\begin{cases}
\pi_0 + \pi_1 + \pi_2 + \pi_3 = 1 
\\ \pi_0 = \pi_0 * \frac{9} {10} + \pi_1 * \frac{9} {10} + \pi_2 * \frac{8} {10} + \pi_3 * \frac{9}{10} 
\\ \pi_1 = \pi_0 * \frac{1}{10} + \pi_2 * \frac{1}{10} + \pi_1 * \frac{1}{10} + \pi_3 * \frac{1}{10}
\\ \pi_2 = \pi_1 * \frac{1}{10} + \pi_2 * \frac{1}{10} \\
\pi_3 = \pi_2 * \frac 1 {10} \end{cases}\ \implies 1000 *\pi_3 = 1 \implies \pi_3 = 0.001 $$\\

```{r}
transition_matrix = matrix(c(0.9, 0.1, 0  , 0,
                             0.9, 0  , 0.1, 0, 
                             0.8, 0  , 0.1, 0.1, 
                             0.9, 0.1, 0  , 0   ), nrow=4, ncol=4, byrow=TRUE)
transition_matrix
chain <- new("markovchain", transitionMatrix = transition_matrix)
steadyStates(chain)
```

## 3

3.  Determine approximately the sample size N which guarantees that the absolute error \|pˆn − pn\| of the estimate pˆn is below 0.03 with confidence level of at least 95 percent. Rerun the experiments for n = 1000 with the determined size N to illustrate the confidence interval and confidence level.

```{r}
#th_mu = mean(get_N_results(1e5, 1e3, "003"))
th_mu = 0.6324
#th_sd = sd(get_N_results(1e5, 1e3, "003"))
th_sd = 0.4823
```

Let $c=0.03$, then we have

$$ P(|\hat{p_n} - p_n|< c)$$

Confidence: $$
1 - \alpha = 0.95 \leq P(|\hat p_n - p_n|<c) \\
 P(\frac{|\hat p_n - p_n|\sqrt{N}} {\sigma} \leq \frac {c\sqrt{N}} {\sigma} ) \approx 1- 2Ф(-\frac {\sqrt{N}} \sigma 0.03)  \\
Ф(-\frac {\sqrt{N}} \sigma 0.03) = \frac \alpha 2 = 0.025 \\
\sigma_{theoretical} = 0.4823 \\
-\frac {\sqrt{N}} \sigma * 0,03 \leq Ф^{-1}(0.025) \\
-\frac {\sqrt{N}} {0.4823} * 0,03 \leq -1.96 \\
\sqrt{N} \gtrapprox 31.5 \\
N_{min} \approx 990
$$\

```{r}
error_prob <- function(n, N, theoretical, eps, num_iter){
  ans = 0
  for (i in 1:num_iter){
    curr_res = get_N_results(N, n, "003")
    if (abs(mean(curr_res) - theoretical)< eps){
      ans = ans + 1
    }
  }
  return (ans/num_iter)
}
```

```{r}
inter = 0.03
alpha = 0.95
th_min_N = (-qnorm((1-alpha)/2)*th_sd/inter)**2
th_min_N = 990
error_prob(1000, th_min_N, th_mu, 0.03, 456)
```

# Problem 2

In the setting of Problem 1, assume that the random digit generation stops at the first occurrence of the TLN (i.e., that the state S4 of the Markov chain is now absorbing). In this problem, you will estimate the average length of such sequences (i.e., the average time till absorption in the Markov chain).

## 1

Make necessary amendments to the transition probabilities matrix P above and solve the corresponding system to find the expected time E(T ) till absorption

The state s3 is absorbing. The transition matrix will be:

$$P_{new}=\begin{pmatrix}
0.9 & 0.1 & 0 & 0
\\0.9 & 0 & 0.1 & 0
\\0.8 & 0 & 0.1 & 0.1
\\0 & 0 & 0 & 1
\end{pmatrix}$$\

System of equations: $$\begin{cases}
\mu_3 = 0
\\ \mu_0 = 1+ \mu_0 * \frac{9} {10} + \mu_1 * \frac{1} {10}
\\ \mu_1 = 1 + \mu_0 * \frac{9}{10} + \mu_2 * \frac{1}{10}
\\ \mu_2 = 1 + \mu_0 * \frac{8}{10} + \mu_2 * \frac{1}{10} + \mu_2 * \frac{1}{10}
\end{cases}\ \implies \mu_0 = 1000 $$

```{r}
transition_matrix = matrix(c(0.9, 0.1, 0  , 0,
                             0.9, 0  , 0.1, 0, 
                             0.8, 0  , 0.1, 0.1, 
                             0  , 0  , 0  , 1   ), nrow=4, ncol=4, byrow=TRUE)
chain <- new("markovchain", transitionMatrix = transition_matrix)
meanAbsorptionTime(chain)
```

As we can see, the expected time for absorbing is 1000.

## 2

Estimate numerically the expected length E(T) till the first occurrence of the TLN by running a sufficiently large number N of experiments

```{r}
 get_first_occur <- function(row_seq, desired=c(0,0,3)){
   for (i in 1:(length(row_seq)-2)){
     if (row_seq[i] == desired[1] && row_seq[i+1] == desired[2] && row_seq[i+2] == desired[3]){
       return(i+2)
     }
   }
   return(0)
 }
```

```{r}
absorb_time <-function(desired=c(0,0,3), cycle_step=1000){
  cycle_num = 0
  row_data = sample(0:9, size=cycle_step, r=T)
   
  while (TRUE){
    result = get_first_occur(row_data, desired)
    if (result != 0){
      return (result + (cycle_step-2)*cycle_num)
    }
    new_data = sample(0:9, size=cycle_step, r=T)
    new_data[1] = row_data[cycle_step-1]
    new_data[2] = row_data[cycle_step]
    row_data = new_data
    cycle_num = cycle_num + 1
  }
}
```

```{r}
get_expected_length <- function(N, cycle_step=1000, desired=c(0,0,3)){
  result = sample(0, size=N, replace = T)
  for (i in 1:N){
    result[i] <- absorb_time()
  }
  return(result)
}
```

```{r}
res <- get_expected_length(1e5)
mean(res)
#get_first_occur(sample(0:9, 10, replace=T), c(0,0,3))
absorb_time(cycle_step=100)
```

## 3

Let $c=10$:

Confidence: \$\$ 1 - \alpha = 0.95 \leq P(\|\hat T_n - T_n\|\<c) \\

P(\|\hat T_n - T_n\|\>c) \leq \frac{c^2}{\sigma^2} = \frac{c^2}{\sigma_{th}^2/N} \\ N \geq \\frac{\\sigma\_{th}\^2}{\\alpha\*c\^2} \\

N\_{min} \approx 200,000 \$\$

```{r}
num_iter = 34
get_sample_means <- function(N_min, num_iter=11, desired="003"){
  results <- sample(0, num_iter, replace = T)
  for (i in 1:num_iter){
    results[i] <- mean(get_expected_length(N_min, desired = desired))
  }
  return (results)
}
```

```{r}
alpha = 0.05
eps=10
th_sd_abs = 1001.7
#th_sd_abs  = sd(get_expected_length(1e7))
th_mu_abs = 999.1
#th_mu_abs  = mean(get_expected_length(1e7))

th_min_N_abs = th_sd_abs*th_sd_abs/(alpha*eps*eps)
th_min_N_abs = floor(th_min_N_abs)
th_min_N_abs
```

### Validate results

```{r}
num_iter = 11
res = get_sample_means(th_min_N_abs, num_iter)
print(res)
mean(abs(res-th_mu_abs)<10)
```


## Exponential distribution

First of all lets generate our variables that we need to use in CI formulas

```{r}
mu = 3/10
M <- 1000
N <- 1000
x <- matrix(rexp(N*M, rate = 1/mu), nrow = N)
sample_mean <- colMeans(x)
sample_sd <- apply(x, 2, sd)
theoretical_sd = mu
```

Using the exact distribution of the statistics 2λnX (show it is χ 2 2n and then use quantiles of the latter to get the interval endpoints)

So lets look how our approach using chi-squared distribution works:

```{r}
for (alpha in c(.01, .05, .1)){
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean((2*N*sample_mean) / qchisq(1-alpha/2, df=N*2) <= mu & mu <= (2*N * sample_mean)/qchisq(alpha/2, df=N*2)), "\n")
  cat("    maximal CI length is", max(2* (2 * N * sample_mean) / quantile(rchisq(2*N, df = 2*N), probs = alpha/2)), "\n")
  cat("    mean CI length is", mean(2* (2 * N * sample_mean) / quantile(rchisq(2*N, df = 2*N), probs = alpha/2)), "\n")
}
```

Using the normal approximation N (µ, σ2 ) for X; the parameters are µ = θ and σ 2 = s 2/n, where s 2 = θ 2 is the population variance (i.e., variance of the original distribution E (λ)). In other words, we form the Z-statistics Z := √ n(X − θ)/θ and use the fact that it is approximately standard normal N (0, 1) to find that P(\|θ − X\| ≤ zβθ/√ n) = P(\|Z\| ≤ zβ) = 2β − 1. in other words, θ is with probability 2β − 1 within X ± zβθ/√ n. 
Let`s see how this approach using the normal approximation works

```{r}
for (alpha in c(.01, .05, .1)){
  beta = 1 - alpha / 2
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(mu - sample_mean) <= qnorm(beta) * theoretical_sd / sqrt(N)), "\n")
  cat("    CI length is", mean(2 * qnorm(beta) * theoretical_sd / sqrt(N)), "\n")
}
```

 The confidence interval constructed above uses the unknown variance s
2 = θ
2 and is of little use in practice.
Instead, we can solve the double inequality
|θ − X| ≤ zβθ/√
n
for θ and get another confidence interval of confidence level 2β −1 that is independent of the unknown parameter.

```{r}
for (alpha in c(.01, .05, .1)){
  beta = 1 - alpha / 2
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(mu - sample_mean) <= qnorm(beta) * sample_sd / sqrt(N)), "\n")
  cat("    CI length is", mean(2 * qnorm(beta) * sample_sd / sqrt(N)), "\n")
}
```

Another (and a more universal approach) to get rid of the dependence on θ in (2) is to estimate s via the sample
standard error and use approximation of X via Student t-distribution

Let`s see how approach using the Student t distribution works:

```{r}
for (alpha in c(.01, .05, .1)){
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(sample_mean-mu) < qt(1-alpha/2, N-1)*sample_sd/sqrt(N)), "\n")
  cat("    maximal CI length is", 2*qt(1-alpha, N-1) * max(sample_sd) / sqrt(N), "\n")
  cat("    mean CI length is", 2*qt(1-alpha/2, N-1) * mean(sample_sd) / sqrt(N), "\n")
}
```

## Poisson distribution

#### Repeat parts (2)–(4) of Problem 3 (with corresponding amendments) for a Poisson distribution P(θ).

How approach using the normal approximation works:

```{r}
mu = 3/10
M <- 2000
N <- 1000
x <- matrix(rpois(N*M, lambda = mu), nrow = N)
sample_mean <- colMeans(x)
theoretical_sd = sqrt(mu)
sample_sd <- apply(x, 2, sd)
for (alpha in c(.01, .05, .1)){
  beta = 1 - alpha / 2
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(mu - sample_mean) <= qnorm(beta) * theoretical_sd / sqrt(N)), "\n")
  cat("    CI length is", max(2 * qnorm(beta) * theoretical_sd / sqrt(N)), "\n")
}
```


```{r}
for (alpha in c(.01, .05, .1)){
  beta = 1 - alpha / 2
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(mu - sample_mean) <= qnorm(beta) * sample_sd / sqrt(N)), "\n")
  cat("    CI length is", mean(2 * qnorm(beta) * sample_sd / sqrt(N)), "\n")
}
```

How approach using the Student t distribution works:

```{r}
for (alpha in c(.01, .05, .1)){
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(sample_mean-mu) < qt(1-alpha/2, N-1)*sample_sd/sqrt(N)), "\n")
  cat("    maximal CI length is", 2*qt(1-alpha, N-1) * max(sample_sd) / sqrt(N), "\n")
  cat("    mean CI length is", 2*qt(1-alpha/2, N-1) * mean(sample_sd) / sqrt(N), "\n")
}
```
