---
title: "Untitled"
output: html_document
date: "2022-11-27"
---

## Exponential distribution

First of all lets generate our variables that we need to use in CI formulas

```{r}
mu = 3/10
M <- 1000
N <- 1000
x <- matrix(rexp(N*M, rate = 1/mu), nrow = N)
sample_mean <- colMeans(x)
sample_sd <- apply(x, 2, sd)
theoretical_sd = mu
```

Using the exact distribution of the statistics 2λnX (show it is χ 2 2n and then use quantiles of the latter to get the interval endpoints)

So lets look how our approach using chi-squared distribution works:

```{r}
for (alpha in c(.01, .05, .1)){
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean((2*N*sample_mean) / qchisq(1-alpha/2, df=N*2) <= mu & mu <= (2*N * sample_mean)/qchisq(alpha/2, df=N*2)), "\n")
  cat("    maximal CI length is", max(2* (2 * N * sample_mean) / quantile(rchisq(2*N, df = 2*N), probs = alpha/2)), "\n")
  cat("    mean CI length is", mean(2* (2 * N * sample_mean) / quantile(rchisq(2*N, df = 2*N), probs = alpha/2)), "\n")
}
```

Using the normal approximation N (µ, σ2 ) for X; the parameters are µ = θ and σ 2 = s 2/n, where s 2 = θ 2 is the population variance (i.e., variance of the original distribution E (λ)). In other words, we form the Z-statistics Z := √ n(X − θ)/θ and use the fact that it is approximately standard normal N (0, 1) to find that P(\|θ − X\| ≤ zβθ/√ n) = P(\|Z\| ≤ zβ) = 2β − 1. in other words, θ is with probability 2β − 1 within X ± zβθ/√ n. 
Let`s see how this approach using the normal approximation works

```{r}
for (alpha in c(.01, .05, .1)){
  beta = 1 - alpha / 2
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(mu - sample_mean) <= qnorm(beta) * theoretical_sd / sqrt(N)), "\n")
  cat("    CI length is", mean(2 * qnorm(beta) * theoretical_sd / sqrt(N)), "\n")
}
```

 The confidence interval constructed above uses the unknown variance s
2 = θ
2 and is of little use in practice.
Instead, we can solve the double inequality
|θ − X| ≤ zβθ/√
n
for θ and get another confidence interval of confidence level 2β −1 that is independent of the unknown parameter.

```{r}
for (alpha in c(.01, .05, .1)){
  beta = 1 - alpha / 2
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(mu - sample_mean) <= qnorm(beta) * sample_sd / sqrt(N)), "\n")
  cat("    CI length is", mean(2 * qnorm(beta) * sample_sd / sqrt(N)), "\n")
}
```

Another (and a more universal approach) to get rid of the dependence on θ in (2) is to estimate s via the sample
standard error and use approximation of X via Student t-distribution

Let`s see how approach using the Student t distribution works:

```{r}
for (alpha in c(.01, .05, .1)){
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(sample_mean-mu) < qt(1-alpha/2, N-1)*sample_sd/sqrt(N)), "\n")
  cat("    maximal CI length is", 2*qt(1-alpha, N-1) * max(sample_sd) / sqrt(N), "\n")
  cat("    mean CI length is", 2*qt(1-alpha/2, N-1) * mean(sample_sd) / sqrt(N), "\n")
}
```

## Poisson distribution

#### Repeat parts (2)–(4) of Problem 3 (with corresponding amendments) for a Poisson distribution P(θ).

How approach using the normal approximation works:

```{r}
mu = 3/10
M <- 2000
N <- 1000
x <- matrix(rpois(N*M, lambda = mu), nrow = N)
sample_mean <- colMeans(x)
theoretical_sd = sqrt(mu)
sample_sd <- apply(x, 2, sd)
for (alpha in c(.01, .05, .1)){
  beta = 1 - alpha / 2
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(mu - sample_mean) <= qnorm(beta) * theoretical_sd / sqrt(N)), "\n")
  cat("    CI length is", max(2 * qnorm(beta) * theoretical_sd / sqrt(N)), "\n")
}
```


```{r}
for (alpha in c(.01, .05, .1)){
  beta = 1 - alpha / 2
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(mu - sample_mean) <= qnorm(beta) * sample_sd / sqrt(N)), "\n")
  cat("    CI length is", mean(2 * qnorm(beta) * sample_sd / sqrt(N)), "\n")
}
```

How approach using the Student t distribution works:

```{r}
for (alpha in c(.01, .05, .1)){
  cat("For confidence level", 1-alpha, "\n")
  cat("    The fraction of CI's containing the parameter is", 
      mean(abs(sample_mean-mu) < qt(1-alpha/2, N-1)*sample_sd/sqrt(N)), "\n")
  cat("    maximal CI length is", 2*qt(1-alpha, N-1) * max(sample_sd) / sqrt(N), "\n")
  cat("    mean CI length is", 2*qt(1-alpha/2, N-1) * mean(sample_sd) / sqrt(N), "\n")
}
```
